\mytitle

### 2015-07-28 MSc_diss.tex
Continued writing text about model validation

\begin{table} % model setup: VOCs
\caption{Model setup: VOCs concentrations (ppb)}\label{tab:setupVOCs}
\centering
\begin{tabular}{ccccccccc}
\hline
Experiment number & $CH_4$ & $C_2H_6$ & $C_3H_8$ & $nC_4H_{10}$ & $iC_4H_{10}$ & $nC_5H_{12}$ & $iC_5H_{12}$ \\
\hline
A  & 1780 &	0.620 &	0.100 &	0.050 &	0.250 &	0.250 &	0.250 \\
B  & 1807 &	1.258 &	0.390 &	0.245 &	0.325 &	0.275 &	0.425 \\
C  & 1834 &	1.896 &	0.680 &	0.440 &	0.400 &	0.300 &	0.600 \\
D  & 1861 &	2.534 &	0.970 &	0.635 &	0.475 &	0.325 &	0.775 \\
E  & 1888 &	3.172 &	1.260 &	0.830 &	0.550 &	0.350 &	0.950 \\
F  & 1915 &	3.810 &	1.550 &	1.025 &	0.625 &	0.375 &	1.125 \\
G  & 1942 &	4.448 &	1.840 &	1.220 &	0.700 &	0.400 &	1.300 \\
H  & 1969 &	5.086 &	2.130 &	1.415 &	0.775 &	0.425 &	1.475 \\
I  & 1996 &	5.724 &	2.420 &	1.610 &	0.850 &	0.450 &	1.650 \\
J  & 2023 &	6.362 &	2.710 &	1.805 &	0.925 &	0.475 &	1.825 \\
K  & 2050 &	7.000 &	3.000 &	2.000 &	1.000 &	0.500 &	2.000 \\
\hline
\end{tabular}
\end{table}
\subsection{Model validation}
Due to imperfection of the chemical mechanism time evolution of species concentrations given by the model does not completely agree with the photochemical theory, but reproduces it to some extent. Figure \ref{fig:model_validation} shows an example of output from the model, namely the changes in mixing ratios of the main species within 24-hours of simulation time under contrasting $NO_x$ and $VOC$ conditions. The values shown in Figure \ref{fig:model_validation} have been justified from the theoretical point of view and also compared with those derived from the Atmospheric Chemistry Learner's Box Model (ACLBM, http://guille.zicer.uea.ac.uk/ACLBM). ACLBM is an on-line box model for atmospheric chemistry developed by Prof. Roland von Glasow (University of East Anglia) that allows the selection of several chemical reaction mechanisms, choice of initial concentrations and emission fluxes. For comparison with our modelling data the ACLBM was set up and initialized using values from Tables \ref{tab:setupO3CO}, \ref{tab:setupNOx} and \ref{tab:setupVOCs}, and the '$O_3$-$NO_x$-$CO$-$CH_4$' chemical mechanism was chosen since it is the closest version of the chemical mechanism available in the ACLBM to our model.

In theory at low $NO_x$ levels a net $O_3$ removal takes place because there is not enough $NO$ around that could react with the peroxy radicals ($HO_2$ and $RO_2$) to produce $NO_2$, which photolysis in conjunction with the reaction (\ref{reac:O3P+O2}) is the main source of $O_3$ in the troposphere. Some additional $O_3$ removal occurs because $HO_2$ radicals can react with $O_3$ leading to the regeneration of $OH$ radicals as part of an $O_3$ depleting $OH$-$HO_2$ ($HO_x$) cycle. In other words, the following reaction sequence prevails:
\begin{equation}\label{reac:O3+hv}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation}\label{reac:O1D+H2O}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation}\label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P
\end{equation}
\begin{equation}\label{reac:O3P+O2}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
\begin{equation}\label{reac:OH+CO}
OH + CO \rightarrow HO_2 + CO_2
\end{equation}
\begin{equation}\label{reac:OH+RH}
OH + RH \rightarrow RO_2 + H_2O
\end{equation}
\begin{equation}\label{reac:HO2+HO2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation}\label{reac:RO2+HO2}
RO_2 + HO_2 \rightarrow ROOH + O_2
\end{equation}
\begin{equation}\label{reac:O3+HO2}
O_3 + HO_2 \rightarrow OH + 2O_2
\end{equation}
In this case our modelling data agrees well with the theory (Figure \ref{fig:O3valid}), because $O_3$ mixing ratio decreases with time, and this decrease is comparable to one derived from ACLBM (by 6 ppb and 4.5 ppb at low and high $VOCs$ concentrations respectively).

The concentration of $NO_x$ is the main factor which determines whether $O_3$ is produced or removed in the atmosphere \citep{Fowler2008}. Theoretically there are three regimes of $O_3$ production that depend on $NO_x$ levels: low $NO_x$ is characterized by net $O_3$ removal (as seen previously), intermediate $NO_x$ - by net $O_3$ formation, and high $NO_x$ - by net $O_3$ formation or removal depending on $VOCs$ concentration. Since the chemical mechanism of our model does not include reactions of nitric acid ($HNO_3$) we won't expect to see a good representation of high $NO_x$ regime in our model. Therefore, further we consider just a contrasting case.

At higher $NO_x$ levels, reactions of the peroxy radicals with $NO$ provide a link required for catalytic generation of $O_3$ via
\begin{equation}\label{reac:HO2+NO}
HO_2 + NO \rightarrow OH + NO_2
\end{equation}
\begin{equation}\label{reac:RO2+NO}
RO_2 + NO \rightarrow RO + NO_2
\end{equation}
\begin{equation}\label{eq:NO2+hv}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation} %same
!!!!! same O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
The competition between reactions (\ref{reac:HO2+NO}) and (\ref{reac:HO2+HO2}) for $HO_2$, and reactions (\ref{reac:RO2+NO}) and (\ref{reac:RO2+HO2}) for $RO_2$, allows a greater number of free-radical propagated $O_3$ forming cycles to happen prior to termination, therefore resulting in a net $O_3$ formation (as it seen in Figure \ref{fig:O3valid}). It appears that in this case $O_3$ formation is somewhat sensitive to changes in the concentrations of $VOCs$ as well.

Under the prevailing conditions, NOx levels remain sufficiently low that peroxide formation via... continues to be the major radical sink. Ozone formation in this regime is often referred to as 'NOx limited' or 'NOx sensitive', because the O3 formation rate increases with increasing NOx concentrations.

@@

The concentration of NOx is the main factor which determines whether O3 is produced or removed in the atmosphere.
changing balance of the free-radical reactions, and their impact on O3, as the level of NOx increases.

NO and NO2 exist in a state equilibrium with rapid interconversion between the two molecules in the troposphere.

@@

\newpage
\clearpage
\thispagestyle{empty}% empty page style (?)
\input{diss_valid_vs_time}

\section{Results} \label{sec:res}
\begin{figure}[h] % Sillman1999
\includegraphics[width=0.5\linewidth]{{./pics/Sillman1999_Fig1_adapted}.png}
\caption{Isopleths giving net rate of ozone production (ppb/h, solid lines) as a function of $VOC$ (ppbC) and $NO_x$ (ppb) for mean summer daytime meteorology and clear skies. The solid lines represent production rates of 1, 2.5, 5, 10, 15, 20 and 30 ppb/h. Adapted from \citep{Sillman1999}.}\label{fig:Sillman1999}
\end{figure}
\begin{figure}[h] % allAN_netO3_rate
\includegraphics[width=\linewidth]{D:/FACSIMILE/ANsCBmodel/matlab/ANsCB_pics/ANsNOxVOC/allAN/chem_allAN_netO3_rate.png}
\caption{allAN, netO3 rate}\label{fig:allAN_netO3_rate}
\end{figure}

>>>>

\begin{table} % model setup: VOCs
\caption{Model setup: VOCs concentrations (ppb)}\label{tab:setupVOCs}
\centering
\begin{tabular}{ccccccccc}
\hline
Experiment number & $CH_4$ & $C_2H_6$ & $C_3H_8$ & $nC_4H_{10}$ & $iC_4H_{10}$ & $nC_5H_{12}$ & $iC_5H_{12}$ \\
\hline
A  & 1780 &	0.620 &	0.100 &	0.050 &	0.250 &	0.250 &	0.250 \\
B  & 1807 &	1.258 &	0.390 &	0.245 &	0.325 &	0.275 &	0.425 \\
C  & 1834 &	1.896 &	0.680 &	0.440 &	0.400 &	0.300 &	0.600 \\
D  & 1861 &	2.534 &	0.970 &	0.635 &	0.475 &	0.325 &	0.775 \\
E  & 1888 &	3.172 &	1.260 &	0.830 &	0.550 &	0.350 &	0.950 \\
F  & 1915 &	3.810 &	1.550 &	1.025 &	0.625 &	0.375 &	1.125 \\
G  & 1942 &	4.448 &	1.840 &	1.220 &	0.700 &	0.400 &	1.300 \\
H  & 1969 &	5.086 &	2.130 &	1.415 &	0.775 &	0.425 &	1.475 \\
I  & 1996 &	5.724 &	2.420 &	1.610 &	0.850 &	0.450 &	1.650 \\
J  & 2023 &	6.362 &	2.710 &	1.805 &	0.925 &	0.475 &	1.825 \\
K  & 2050 &	7.000 &	3.000 &	2.000 &	1.000 &	0.500 &	2.000 \\
\hline
\end{tabular}
\end{table}
\subsection{Model performance}
Due to imperfection of the chemical mechanism time evolution of species concentrations given by the model does not completely agree with the photochemical theory, but reproduces it to some extent. Figure \ref{fig:model_validation} shows an example of output from the model, namely the changes in mixing ratios of the main species within 24-hours of simulation time under contrasting $NO_x$ and $VOCs$ conditions. The values shown in Figure \ref{fig:model_validation} have been justified from the theoretical point of view and also compared with those derived from the Atmospheric Chemistry Learner's Box Model (ACLBM, http://guille.zicer.uea.ac.uk/ACLBM). ACLBM is an on-line box model for atmospheric chemistry developed by Prof. Roland von Glasow (University of East Anglia) that allows the selection of several chemical reaction mechanisms, choice of initial concentrations and emission fluxes. For comparison with our modelling data the ACLBM was set up and initialized with values from Tables \ref{tab:setupO3CO}, \ref{tab:setupNOx} and \ref{tab:setupVOCs}, and the '$O_3$-$NO_x$-$CO$-$CH_4$' chemical mechanism was chosen as it is the closest version of the chemical mechanism available in the ACLBM to our model.

The concentration of $NO_x$ is the main factor which determines whether $O_3$ is produced or removed in the atmosphere, and theoretically there are three regimes of $O_3$ production that depend on $NO_x$ levels: at low $NO_x$ levels the regime is characterized by net $O_3$ removal, at intermediate $NO_x$ (also called $NO_x$-sensitive) - by net $O_3$ formation, and at high $NO_x$ - by net $O_3$ formation or removal depending more on $VOCs$ concentration (therefore called $VOC$-sensitive) (e.g., \citep{Fowler2008}). When the prevailing $NO_x$ levels are low, a net $O_3$ removal takes place because there is not enough $NO$ around to react with the peroxy radicals ($HO_2$ and $RO_2$) to produce $NO_2$, which photolysis in conjunction with the reaction (\ref{reac:O3P+O2}) is the main source of $O_3$ in the troposphere. Some additional $O_3$ removal in this regime occurs due to reaction of $HO_2$ radicals with $O_3$ leading to the regeneration of $OH$ radicals as part of an $O_3$ depleting $OH$-$HO_2$ ($HO_x$) cycle. In other words, the following reaction sequence predominates:
\begin{equation}\label{reac:O3+hv}
O_3 + hv \rightarrow O(^1D) + O_2
\end{equation}
\begin{equation}\label{reac:O1D+H2O}
O(^1D) + H_2O \rightarrow OH + OH
\end{equation}
\begin{equation}\label{reac:O1D=O3P}
O(^1D) \xrightarrow{M} O(^3P
\end{equation}
\begin{equation}\label{reac:O3P+O2}
O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
\begin{equation}\label{reac:OH+CO}
OH + CO \rightarrow HO_2 + CO_2
\end{equation}
\begin{equation}\label{reac:OH+RH}
OH + RH \rightarrow RO_2 + H_2O
\end{equation}
\begin{equation}\label{reac:HO2+HO2}
HO_2 + HO_2 \rightarrow H_2O_2 + O_2
\end{equation}
\begin{equation}\label{reac:RO2+HO2}
RO_2 + HO_2 \rightarrow ROOH + O_2
\end{equation}
\begin{equation}\label{reac:O3+HO2}
O_3 + HO_2 \rightarrow OH + 2O_2
\end{equation}
At low $NO_x$ regime our modelling data agrees well with the theory (Figure \ref{fig:O3valid}). $O_3$ mixing ratio decreases with time, and this decrease is comparable to one derived from ACLBM (by 6 ppb and 4.5 ppb at low and high $VOCs$ concentrations respectively). $NO$ is rapidly converted to $NO_2$ representing a state equilibrium between the two molecules in the troposphere (Figures \ref{fig:NOvalid}, \ref{fig:NO2valid}). $CO$ is lost due to its reaction with $OH$ (Figure \ref{fig:COvalid}). Mixing ratios of other species in our model are of the order of magnitude as in ACLBM.

\input{diss_valid_vs_time}

When the prevailing $NO_x$ levels are more than 0.5 ppb \citep{Sillman1999}, but less than certain critical value, reactions (\ref{reac:HO2+NO}) and (\ref{reac:RO2+NO}) represent the dominant reaction pathways for $HO_2$ and $RO_2$ providing a link for catalytic generation of $O_3$:
\begin{equation}\label{reac:HO2+NO}
HO_2 + NO \rightarrow OH + NO_2
\end{equation}
\begin{equation}\label{reac:RO2+NO}
RO_2 + NO \rightarrow RO + NO_2
\end{equation}
\begin{equation}\label{eq:NO2+hv}
NO_2 + hv \rightarrow NO + O(^3P)
\end{equation}
\begin{equation} %same
!!!!! same O(^3P) + O_2 \xrightarrow{M} O_3
\end{equation}
The competition between reactions (\ref{reac:HO2+NO}) and (\ref{reac:HO2+HO2}) for $HO_2$, and reactions (\ref{reac:RO2+NO}) and (\ref{reac:RO2+HO2}) for $RO_2$, allows a greater number of free-radical propagated $O_3$ forming cycles to happen prior to termination (reactions (\ref{reac:HO2+HO2}) and (\ref{reac:RO2+HO2})), therefore resulting in a net $O_3$ formation.

When the prevailing $NO_x$ levels are high, reactions (\ref{reac:HO2+NO}) and (\ref{reac:RO2+NO}) dominate for $HO_2$ and $RO_2$, making the rate of $O_3$ production being controlled by the availability of these radicals. The availability of $OH$ radical is also important at high $NO_x$ levels because $OH$ connects the $O_3$ production rate with rates of its reactions with $CO$ and $RH$ (reactions (\ref{reac:OH+CO}) and (\ref{reac:OH+RH})), therefore making this regime $VOC$-sensitive. Plus at this point the reaction of $OH$ with $NO_2$ to form nitric acid ($HNO_3$) becomes the major termination reaction for free radicals:
\begin{equation}\label{reac:OH+NO2}
OH + NO_2 \xrightarrow{M} HNO_3
\end{equation}
Under these circumstances, the number of free-radical propagated $O_3$ forming cycles decreases with increasing $NO_x$, and the production rate of $O_3$ decreases. However, if the concentrations of $VOCs$ are high, reactions of $OH$ with $CO$ and $RH$ ((\ref{reac:OH+CO}), (\ref{reac:OH+RH})) can compete with reaction (\ref{reac:OH+NO2}), leading to an increase in the $O_3$ production rate.

Since the chemical mechanism of our model does not include the reaction of $HNO_3$ formation, a good representation of $VOC$-sensitive regime by our model is not expected. It also makes it difficult to split $NO_x$-sensitive and $VOC$-sensitive regimes in our model, because this split is determined by the recurrence (size) of the peroxide- and nitric-acid forming reactions \citep{Sillman1999}, while our mechanism lacks one of these reactions. However, needing to evaluate performance of our model under higher $NO_x$ conditions anyway below we consider just a contrasting case of $NO_x$ levels.

In our model a net $O_3$ formation first occurs at 250 ppt of $NO_x$ and remains so at 10 ppb of $NO_x$ too (Figure \ref{fig:noAN_netO3_rate}).

\begin{figure}[h] % noAN_netO3rate_withINORG_ppbC
\includegraphics[width=0.5\linewidth]{D:/FACSIMILE/ANsCBmodel/matlab/ANsCB_pics/ANsNOxVOC/noAN/chem_noAN_netO3rate_withINORG_ppbC.png}
\caption{noAN, netO3 rate}\label{fig:noAN_netO3_rate}
\end{figure}

\begin{figure}[h] % Sillman1999
\includegraphics[width=0.5\linewidth]{{./pics/Sillman1999_Fig1_adapted}.png}
\caption{Isopleths giving net rate of ozone production (ppb/h, solid lines) as a function of $VOC$ (ppbC) and $NO_x$ (ppb) for mean summer daytime meteorology and clear skies. The solid lines represent production rates of 1, 2.5, 5, 10, 15, 20 and 30 ppb/h. Adapted from \citep{Sillman1999}.}\label{fig:Sillman1999}
\end{figure}

It appears that in this case $O_3$ formation is somewhat sensitive to changes in the concentrations of $VOCs$ as well.

@@

@@

\section{Results} \label{sec:res}
\begin{figure}[h] % allAN_netO3_rate
\includegraphics[width=\linewidth]{D:/FACSIMILE/ANsCBmodel/matlab/ANsCB_pics/ANsNOxVOC/allAN/chem_allAN_netO3_rate.png}
\caption{allAN, netO3 rate}\label{fig:allAN_netO3_rate}
\end{figure}
